```{r}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will practicing with `dpylr` functions using `starwars` dataset:

```{r}
starwars
head(starwars)
?starwars
```

## Subset rows with filter

`filter()` allows us to subset the rows of a datset based on the values

```{r}
## Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Human")
humans 
filter(humans, sex == "male")

## filters multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

We can use different comparison operators to filter:

-   `!` is "not"

-   `==` is "equal:

-   `&` is "and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   i`s.na()` returns TRUE if value is NA

```{r}
## returns characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are not human?
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") # returns false because "grey" is not in vector of colors

filter(starwars, hair_color %in% c("brown", "blond"))
```

`grepl("characters_of_interest", variable_name)` returns TRUE if the value of variable_name contains the characters_of \_interest.

```{r}
filter(starwars, grepl("blond", hair_color))

filter(starwars, grepl("blue", eye_color))
```

You can use `<, >, <=, >=` to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D") # ranking is based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") # ranking is based on levels of the ordered factor cut

"D" < "d" #false
"d" < "D" #true
```

between(x, left, right) is a shorthand for x \>= left & \<= right

```{r}
between(5,1,7) # true because 5 is less than or equal to 7 and greater than or equal to 1
filter(starwars, between(mass, 50, 100))
```

### Practice

Find all star wars characters that:

1.  Are missing a hair color

    ```{r}
    filter(starwars, is.na(hair_color))
    ```

2.  Are shorter than 100cm and are not droids

    ```{r}
    filter(starwars, species != "Droid", height < 100)
    ```

3.  Were born between the years 100 and 200

    ```{r}
    filter(starwars, between(birth_year, 100, 200))
    ```

4.  Your choice

    ```{r}
    filter(starwars, mass >= 100, height >= 180)
    ```

## Order Rows with Arrange

arrange() reorders rows based on the values of column(s)

```{r}
## order by character names
arrange(starwars, name)
# order of character values: aAbBcCdD...

## order by species and birth year
arrange(starwars, species, birth_year)

## use desc() to get birth years in descending order
arrange(starwars, species, desc(birth_year))
```

### Practice

1.  Arrange the starwars dataset from tallest to shortest characters

    ```{r}
    arrange(starwars, desc(height))
    ```

2.  Arrange the starwars dataset from shortest to tallest by species

    ```{r}
    arrange(starwars, species, height) # order by height, tiebraker is alphabetic species

    arrange(starwars, species, height)
    ```

3.  Alphabetize the Star Wars characters by their hair color then eye color

    ```{r}
    arrange(starwars, hair_color, eye_color)
    ```

## Subset Columns with select()

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_color <- select(starwars, hair_color, skin_color, eye_color)
```

We can also select multiple columns in a row

```{r}
# all columns between name and eye_color (inclusive)
select(starwars, name:eye_color)
```

We can also remove columns by name using a -

```{r}
starwars
select(starwars, -films, -vehicles)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

### Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # put species first, then all the other columns

select(starwars, species:sex, everything())

## alphabetize columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

starts_with("abc") matches column names that start with "abc"

ends_with("abc") matches column names that end with "abc"

contains("abc") matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

### Practice

1.  Select each character's homeworld and species. What question might this subset of variables help answer?

    ```{r}
    select(starwars, homeworld, species)
    # this subset of variables could answer the qustion "where do species live?"
    ```

2.  Brainstorm as many ways as possible to select `hair_color, skin_color, eye_color, and birth_year`

    ```{r}
    select(starwars, hair_color:birth_year)
    select(starwars, contains("color"), birth_year)
    select(starwars, contains("_"), birth_year)
    select(starwars, hair_color, skin_color, eye_color, birth_year)
    select(starwars, -c(name:mass, sex:starships))
    ```

## Add new variables with mutate

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

#only keep new variables
transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

```

Summary statistics give one value for the entire column: sum(), mean()

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group by a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm = T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.character(mass))
```

### Practice

1.  Convert mass to pounds

    ```{r}
    mutate(starwars_small,
           weight = mass*2.2)
    ```

## Summarize data with summarize

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#Group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create.

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

## group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species,sex))
```
